Constructing ordinal outcomes for clinical trials

### 2025-06-19 constructing_ordinal_outcomes.tex
Attempt to tighten up intro w Kate+Chris feedback

\begin{abstract}
  Ordinal outcomes (or ``endpoints'') are an increasingly popular choice in
  clinical trials.
  This paper is a guide to techniques for constructing ordinal
  outcomes, aimed at clinicians and their statistical collaborators.
  We categorize and illustrate ways of combining existing binary,
  ordinal, or continuous
  outcomes through three broad mechanisms: sequential composition
  (overriding), hierarchical composition (tie-breaking), and parallel
  composition.
  We discuss how this fine-graining impacts validity and efficiency, and also
  suggest reasons why coarser outcomes might be preferred.
  We also situate ``generalized pairwise comparisons'' and
  ``prioritized outcomes'' approaches, like the win
  ratio and desirability of outcome
  ranking (DOOR) methodologies, in the broader context of ordinal
  outcomes, suggesting that methods and approaches from each field
  might be useful in the other.
\end{abstract}

\tableofcontents
\newpage

@@

\subsection{Ordinal outcomes and their advantages}

Ordinal outcomes (or ``endpoints'') have become increasingly
prominent in clinical
resarch, particularly following the COVID-19 pandemic, when ordinal
scales developed by the WHO and NIAID were used in prominent trials
\citep{beigelRemdesivirTreatmentCovid192020,
marshallMinimalCommonOutcome2020}. They have also
been used in a number of adaptive platform trials
\citep{angusREMAPCAPRandomizedEmbedded2020, walkerCodesigningNovelOrdinal2025}.

Ordinal outcomes rank patient outcomes from better to worse, but
\emph{differences} in patient outcomes are
not treated as meaningful. Binary
outcomes are ordinal -- they are the simplest kind of
ordinal outcome -- and so too are continuous outcomes, though
``treating them ordinally'' involves ignoring their numeric values.

Two desirable features of outcomes in general are \emph{validity} and
\emph{efficiency}.

\textbf{Valid} outcomes reflect
patient benefit and harm: better/worse outcomes are judged respectively
better/worse by patients, are objectively respectively better/worse for
patients, or are reliable surrogates/predictors of outcomes that are
respectively better/worse for patients.

\begin{itemize}
  \item Allow for the incorporation of intercurrent events into one's primary
    outcome, avoiding complexities of alternative strategies like
    principal stratum
    estimands (particular for truncating events like death) while yielding more
    statistical efficiency than traditional binary composite outcomes

  \item Allow for the easy incorporation of safety events into a
    primary outcome, accounting for many different aspects
    of benefit/harm; at the same time, ordinal outcomes can be
    analyzed without a
    fully-specified utility function, which may require more
    extensive development and validation or may vary widely across patients.

  \item Allow for the combination of discrete and continous outcome
    measures (for
      example a numerical quality of life scale with death as the
      overriding worst
    outcome) and allow for the combination of longitudinal and timepoint or
    ``state'' outcomes in a single measure.
\end{itemize}

\textbf{Efficient} outcomes
capture sufficient information to discriminate among better- and worse-off
patients without large sample sizes.

@@

A key feature of ordinal outcomes is the ease with which they can be
\emph{combined}.

\subsection{Desiderata for (ordinal) outcomes: validity and efficiency}

We begin with some general desiderata for outcomes and effect measures in
clinical studies.

@@

An example of an outcome which is efficient but which may be invalid
might be continuously-measured concentrations of a biomarker which is a poor
surrogate for disease progression.

\subsection{The iterative process of constructing and validating}

We want to distinguish \emph{choosing} or \emph{constructing} an
outcome/effect
measure from \emph{validating} the outcome/effect measure using
external data.
This paper is about the first task, choosing (or constructing)
the outcome and
effect measure. In practice choosing an outcome and effect measure and
validating that outcome and effect measure in pre-trial data will be an
iterative process. Our focus in this paper is on the ``positive''
side of the
process--namely coming up with, and tweaking, proposed outcomes and effect
measures--rather than the ``negative'' side, namely testing or checking
proposals with data to assess whether they have desired properties.

@@

It is important to distinguish ``outcomes'' and ``effect measures'' -- both
sometimes referred to under the banner of ``endpoints''. Here we
focus on outcomes; we discuss endpoints in a separate paper \todo{cite}.

@@

@@

\begin{figure}
  \includegraphics[width=7in]{passport_ordinal_outcome_sequential.pdf}
  % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
  \caption{\textbf{Sequential composition (overriding) illustrated
    with the PASSPORT outcome} Ordinal outcome used in the planned
    PASSPORT adaptive
    platform trial. This illustrates sequential composition/overriding:
    of days hospitalized by maximum pSOFA while hospitalized, and of
  both of these by day 28 status.}
  \label{fig:passport_outcome}
\end{figure}

\begin{figure}
  \includegraphics[width=6in]{rajkumar_et_al_outcome.png}
  \caption{\textbf{Hierarchical composition (complete tie-breaking)
      illustrated with the
    the ORBITA-2 outcome.} ``Angina
    Symptom Score'' ordinal outcome develped in
    \citep{rajkumarPlaceboControlledTrialPercutaneous2023}. This illustrates
    (1) hierarchical composition or complete tie-breaking (of units
      of antianginal medication by
    episodes of angina) and also (2) overriding (of angina episodes and
  angina medication by adverse clinical events)}
  \label{fig:rajkumar_et_al_outcome}
\end{figure}

@@

Also, we "chunk" longer hospital stays more aggressively to avoid
small categories, for interpretability's sake.

@@

One point of view is that one would ``ideally'' analyze ordinal outcomes by
reporting the associated expected utilities \todo{cite Berry
  talk; example of
utility-weighted Rankin scale} and that since existing ordinal
methods supply
such utilities implicitly why not just supply them explicitly. A
counterpoint
is that the methods do not, in fact, supply utilities implicitly.
Another is that ordinal outcomes let one ``hedge'' about precise
utillities and this is OK.

>>>>

\begin{abstract}
  Ordinal outcomes are an increasingly popular choice for
  clinical trials.
  This paper is a guide to techniques for constructing ordinal
  outcomes out of pre-existing components, aimed at clinicians and their
  statistical collaborators.
  We discuss three broad mechanisms for combining existing binary,
  ordinal, or continuous
  outcomes into an ordinal composite---overriding (or
  ``sequential composition''), tie-breaking (or ``hierarchical
  composition''), and what we term ``parallel
  composition''---and illustrate each mechanism with a real-world example.
  We discuss how these mechanisms can improve the validity and
  efficiency of ordinal outcomes, but also why coarsening outcomes by
  combining levels might be preferred in practice.
  Our approach is closely related to the literature on prioritized
  outcomes (like win methods and
    desirability of outcome
  ranking methodologies), but we focus on the task of
  constructing ordinal
  outcomes and distinguish this from the related tasks of choosing
  effect measures and estimators and
  handling censoring.
\end{abstract}
\newpage
\tableofcontents
\newpage

@@

\subsection{Ordinal outcomes, their recent popularity and advantages}

Ordinal outcomes (or ``endpoints'') have become increasingly
prominent in clinical
resarch, particularly following the COVID-19 pandemic, when ordinal
scales developed by the WHO and NIAID were used in prominent trials
\citep{beigelRemdesivirTreatmentCovid192020,
marshallMinimalCommonOutcome2020}. Ordinal outcomes have also
been used in a number of adaptive platform trials
\citep{angusREMAPCAPRandomizedEmbedded2020,
walkerCodesigningNovelOrdinal2025}.

Ordinal outcomes rank patient outcomes from better to worse, but
\emph{differences} in patient outcomes are
not treated as meaningful.

Two desirable features of outcomes in general are \emph{validity} and
\emph{efficiency}.
\emph{Valid} outcomes reflect
patient benefit and harm; \emph{efficient} outcomes
are able to discriminate among better- and worse-off
patients without large sample sizes.

A great strength of ordinal outcomes is the ease with which they can
be combined.

This can help with validity by

\begin{itemize}
  \item \emph{Handle intercurrent events} by incorporating them into a primary
    outcome, avoiding complexities of alternative strategies like
    principal stratum
    estimands (particular for truncating events like death) while
    yielding more
    statistical efficiency than traditional binary composite outcomes.

  \item \emph{Jointly summarizing efficacy and safety} by incorporating
    safety events into a
    primary outcome.

  \item \emph{Combining outcome measures of different types} Allow
    for the combination of discrete and continous outcome
    measures (for
      example a numerical quality of life scale with death as the
      overriding worst
    outcome) and allow for the combination of longitudinal and timepoint or
    ``state'' outcomes in a single measure.
\end{itemize}

Moreover it can help with efficiency by

@@

@@

An example of an outcome which is efficient but which may be invalid
might be continuously-measured concentrations of a biomarker which is a poor
surrogate for disease progression, or a continuous outcome that
excludes mortality.

@@

\subsubsection{Non-objectives}

\paragraph{Outcomes versus effect measures} It is important to
distinguish ``outcomes'' and ``effect measures'' -- both
sometimes referred to under the banner of ``endpoints''. Here we
focus on outcomes; we discuss endpoints in a separate paper \todo{cite}.

\paragraph{Constructing versus validating} We want to distinguish
\emph{choosing} or \emph{constructing} an
outcome/effect
measure from \emph{validating} the outcome/effect measure using
external data.
This paper is about the first task, choosing (or constructing)
the outcome and
effect measure. In practice choosing an outcome and effect measure and
validating that outcome and effect measure in pre-trial data will be an
iterative process. Our focus in this paper is on the ``positive''
side of the
process--namely coming up with, and tweaking, proposed outcomes and effect
measures--rather than the ``negative'' side, namely testing or checking
proposals with data to assess whether they have desired properties.

@@

The process of combining ordinal outcomes is more flexible than it
might seem, because many different kinds of outcomes can be treated as ordinal
or summarized in ordinal form. Binary
outcomes are ordinal -- they are the simplest kind of
ordinal outcome -- and so too are continuous outcomes, though
``treating them ordinally'' involves ignoring their numeric values.
Longitudinal trajectory-outcomes can be summarized ordinally---for
example ``worst measurement in the first 28 days of
hospitalization''---and combined with ``state'' outcomes.

@@

\begin{figure}
  \includegraphics[width=7in]{passport_ordinal_outcome_sequential.pdf}
  % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
  \caption{\textbf{Overriding (sequential composition) illustrated
    with the PASSPORT outcome} Ordinal outcome used in the planned
    PASSPORT adaptive
    platform trial. This illustrates sequential composition/overriding:
    of days hospitalized by maximum pSOFA while hospitalized, and of
  both of these by day 28 status.}
  \label{fig:passport_outcome}
\end{figure}

\begin{figure}
  \includegraphics[width=6in]{rajkumar_et_al_outcome.png}
  \caption{\textbf{Tie-breaking (hierarchical composition)
      illustrated with the
    the ORBITA-2 outcome.} ``Angina
    Symptom Score'' ordinal outcome develped in
    \citep{rajkumarPlaceboControlledTrialPercutaneous2023}. This illustrates
    (1) hierarchical composition or complete tie-breaking (of units
      of antianginal medication by
    episodes of angina) and also (2) overriding (of angina episodes and
  angina medication by adverse clinical events)}
  \label{fig:rajkumar_et_al_outcome}
\end{figure}

@@

Also, we "chunk" longer hospital stays more aggressively to avoid
small categories, for the sake of interpretability.

@@

Our approach is closely related to that of
\citet{follmannAnalysisOrderedComposite2020},
who explicitly consider the ordinal outcome implicitly constructed by
win methods from a time-to-event analysis of ordered events, and show
how right-censoring of event times induces interval censoring of the
cosntructed outcome.

One point of view is that one would ``ideally'' analyze ordinal outcomes by
reporting the associated expected utilities \todo{cite Berry
  talk; example of
utility-weighted Rankin scale} and that since existing ordinal
methods supply
such utilities implicitly why not just supply them explicitly. A
counterpoint
is that the methods do not, in fact, supply utilities implicitly.
Another is that ordinal outcomes let one ``hedge'' about precise
utillities and this is OK:
fully-specified utility function may require more
extensive development and validation or may vary widely across patients.
